**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**

### 2021-05-28 Manuscripts/Manuscript by chapters -v1/02-Challenges.Rmd
Updated, added LMEM and rm-ANOVA models in text and in appendix, corrected cubic spline to thin plate spline, changed to default in gam models instead of gp smooths

  ## repeated measures ANOVA in R
#time and treatment interaction model, compound symmetry required by the model
  fit_lme <- lme(y ~ treatment + time + treatment:time,
                 data = dat,
                 random = ~ 1 | subject,
                 correlation = corCompSymm(form = ~ 1 | subject)
  )

@@

  #add model predictions to the dataframe that has the simulated data
  dat$y_pred <- predict(fit_lme)

  #return everything in a list
  return(list(
    dat = dat,
    pred_dat = pred_dat,
    fit_lme = fit_lme

@@

   #plot the model predictions
  p4 <- ggplot(sim_dat$dat, 
               aes(x = time, 
                   y = y, 
                   color = treatment)) +
    geom_point()+
    labs(y='response')+
    geom_line(aes(y = predict(sim_dat$fit_lme), 
                  group = subject, size = "Subjects")) +
    geom_line(data = sim_dat$pred_dat, 
              aes(y = predict(sim_dat$fit_lme, 
                              level = 0, 
                              newdata = sim_dat$pred_dat), 
                  size = "Population")) +
    guides(color = guide_legend(override.aes = list(size = 2)))+
    scale_size_manual(name = "Predictions", 
                      values=c("Subjects" = 0.5, "Population" = 3)) +
    theme_classic() +
    theme(plot.title = element_text(size = 30, 
                                  face = "bold"),
        text=element_text(size=30))+
    thm

  return((p1+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))

>>>>

  ## repeated measures ANOVA in R
#time and treatment interaction model
  fit_anova <- lm(y ~ time + treatment + time * treatment, data = dat)

  #LMEM with compound symmetry

  fit_lme <- lme(y ~ treatment + time + treatment:time,
                 data = dat,
                 random = ~ 1 | subject,
                 correlation = corCompSymm(form = ~ 1 | subject)
  )

@@

  #add model predictions to the dataframe that has the simulated data
  dat$pred_anova <- predict(fit_anova)
  dat$pred_lmem <- predict(fit_lme)

  #return everything in a list
  return(list(
    dat = dat,
    pred_dat = pred_dat,
    fit_lme = fit_lme,
    fit_anova=fit_anova

@@

   #plot the model predictions for rm-ANOVA
  p2 <- ggplot(sim_dat$dat, 
               aes(x = time, 
                   y = y, 
                   color = treatment)) +
    geom_point()+
    labs(y='response')+
    geom_line(aes(y = predict(sim_dat$fit_anova), 
                  group = subject, size = "Subjects")) +
    geom_line(data = sim_dat$pred_dat, 
              aes(y = predict(sim_dat$fit_anova, 
                              level = 0, 
                              newdata = sim_dat$pred_dat), 
                  size = "Population")) +
    guides(color = guide_legend(override.aes = list(size = 2)))+
    scale_size_manual(name = "Predictions", 
                      values=c("Subjects" = 0.5, "Population" = 3)) +
    theme_classic() +
    theme(plot.title = element_text(size = 30, 
                                  face = "bold"),
        text=element_text(size=30))+
    thm

   #plot the model predictions for LMEM
  p4 <- ggplot(sim_dat$dat, 
               aes(x = time, 
                   y = y, 
                   color = treatment)) +
    geom_point()+
    labs(y='response')+
    geom_line(aes(y = predict(sim_dat$fit_lme), 
                  group = subject, size = "Subjects")) +
    geom_line(data = sim_dat$pred_dat, 
              aes(y = predict(sim_dat$fit_lme, 
                              level = 0, 
                              newdata = sim_dat$pred_dat), 
                  size = "Population")) +
    guides(color = guide_legend(override.aes = list(size = 2)))+
    scale_size_manual(name = "Predictions", 
                      values=c("Subjects" = 0.5, "Population" = 3)) +
    theme_classic() +
    theme(plot.title = element_text(size = 30, 
                                  face = "bold"),
        text=element_text(size=30))+
    thm

  return((p1+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))

### 2021-05-28 Manuscripts/Manuscript by chapters -v1/Full_document.tex
Updated, added LMEM and rm-ANOVA models in text and in appendix, corrected cubic spline to thin plate spline, changed to default in gam models instead of gp smooths

\begin{lstlisting}[language=R]
####################Section for calculations######################################
################################################################################

@@

#plotting simulated datapoints and smooth term
p14<-ggplot(data=dat,
            aes(x=time,y=y))+
  geom_point(size=sz+1)+
  scale_color_aaas()+
  labs(y='Simulated \n response')+
  geom_line(data=smooth,
            aes(x=time,
                y=smooth),
            color="#6C581DFF",
            size=sz+1)+
  theme_classic()

#Combining all
b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
  theme(
     text=element_text(size=18)
     )
\end{lstlisting}

@@

\begin{lstlisting}[language=R]
layout1 <- c(
  area(1, 1),
  area( 1, 2),
  area(2, 1),
  area(2, 2),
  area(1, 3, 2)
)

layout2 <- c(
  area(1, 1),
  area( 1, 2),
  area(2, 1),
  area(2, 2),
  area(1, 3, 2),
  area(1,4,2),
  area(1,5,1)
)
\end{lstlisting}

@@

\begin{lstlisting}
## 
## Method: REML   Optimizer: outer newton
## full convergence after 6 iterations.
## Gradient range [-0.0001585015,0.0008415702]
## (score 436.939 & scale 387.386).
## Hessian positive definite, eigenvalue range [0.0001600441,48.99916].
## Model rank =  5 / 5 
## 
## Basis dimension (k) checking results. Low p-value (k-index<1) may
## indicate that k is too low, especially if edf is close to k'.
## 
##        k' edf k-index p-value    
## s(Day)  4   1    0.37  <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{lstlisting}

@@

\begin{lstlisting}
## 
## Family: gaussian 
## Link function: identity 
## 
## Formula:
## StO2_sim ~ s(Day, k = 5, bs = "gp")
## 
## Parametric coefficients:
##             Estimate Std. Error t value Pr(>|t|)    
## (Intercept)   21.578      1.968   10.96   <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Approximate significance of smooth terms:
##          edf Ref.df     F  p-value    
## s(Day) 1.002  1.004 21.54 1.09e-05 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## R-sq.(adj) =  0.172   Deviance explained = 18.1%
## -REML = 436.94  Scale est. = 387.39    n = 100
\end{lstlisting}

@@

\begin{lstlisting}
## 
## Method: REML   Optimizer: outer newton
## full convergence after 8 iterations.
## Gradient range [-0.0001335263,0.001505324]
## (score 415.0769 & scale 254.693).
## Hessian positive definite, eigenvalue range [9.414522e-05,48.49849].
## Model rank =  9 / 9 
## 
## Basis dimension (k) checking results. Low p-value (k-index<1) may
## indicate that k is too low, especially if edf is close to k'.
## 
##                       k' edf k-index p-value    
## s(Day):GroupControl    4   1    0.46  <2e-16 ***
## s(Day):GroupTreatment  4   1    0.46  <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{lstlisting}

@@

\begin{lstlisting}
## 
## Family: gaussian 
## Link function: identity 
## 
## Formula:
## StO2_sim ~ s(Day, by = Group, k = 5, bs = "gp")
## 
## Parametric coefficients:
##             Estimate Std. Error t value Pr(>|t|)    
## (Intercept)   21.578      1.596   13.52   <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Approximate significance of smooth terms:
##                         edf Ref.df      F p-value    
## s(Day):GroupControl   1.004  1.008  1.083   0.299    
## s(Day):GroupTreatment 1.000  1.001 83.768  <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## R-sq.(adj) =  0.456   Deviance explained = 46.7%
## -REML = 415.08  Scale est. = 254.69    n = 100
\end{lstlisting}

@@

gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
            method='REML',
            data  = dat_sim)

gam.check(gam1)
\end{lstlisting}

@@

\begin{lstlisting}
## 
## Method: REML   Optimizer: outer newton
## full convergence after 9 iterations.
## Gradient range [-1.291119e-06,1.637752e-06]
## (score 374.509 & scale 96.64781).
## Hessian positive definite, eigenvalue range [0.02491842,48.04328].
## Model rank =  10 / 10 
## 
## Basis dimension (k) checking results. Low p-value (k-index<1) may
## indicate that k is too low, especially if edf is close to k'.
## 
##                         k'  edf k-index p-value  
## s(Day):GroupControl   4.00 3.84    0.85   0.075 .
## s(Day):GroupTreatment 4.00 1.24    0.85   0.085 .
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{lstlisting}

\begin{lstlisting}[language=R]
summary(gam1)
\end{lstlisting}

\begin{lstlisting}
## 
## Family: gaussian 
## Link function: identity 
## 
## Formula:
## StO2_sim ~ Group + s(Day, by = Group, k = 5, bs = "gp")
## 
## Parametric coefficients:
##                Estimate Std. Error t value Pr(>|t|)    
## (Intercept)      10.503      1.390   7.554 2.87e-11 ***
## GroupTreatment   22.150      1.966  11.266  < 2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Approximate significance of smooth terms:
##                         edf Ref.df       F  p-value    
## s(Day):GroupControl   3.843  3.976   7.975 8.99e-06 ***
## s(Day):GroupTreatment 1.236  1.419 159.701  < 2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## R-sq.(adj) =  0.794   Deviance explained = 80.6%
## -REML = 374.51  Scale est. = 96.648    n = 100
\end{lstlisting}

The resulting model is model \passthrough{\lstinline!gam1!}, which is the model fitted in the main manuscript. By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group . From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\textasciitilde1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 87\% of the variance data.

@@

\begin{lstlisting}
##              df      AIC
## gam_00 3.004277 883.7160
## gam_01 4.008467 842.7602
## gam1   8.395441 750.3446
\end{lstlisting}

@@

##matrix that contains the basis functions evaluated at the points in pdat
    xp <- predict(gam1, newdata = pdat, type = 'lpmatrix')

@@

\begin{lstlisting}[language=R]
#missing data
#create a sequence of 40 random numbers between 1 and 100, these numbers will
#correspond to the row numbers to be randomly erased from the original dataset
missing <- sample(1:100, 40)

@@

\begin{lstlisting}
## # A tibble: 10 x 3
## # Groups:   Day, Group [10]
##      Day Group         n
##    <dbl> <fct>     <int>
##  1     0 Control       2
##  2     0 Treatment     4
##  3     2 Control       7
##  4     2 Treatment     4
##  5     5 Control       2
##  6     5 Treatment     8
##  7     7 Control       5
##  8     7 Treatment     8
##  9    10 Control       4
## 10    10 Treatment     5
\end{lstlisting}

>>>>

\begin{lstlisting}[language=R]
set.seed(1)
####################Section for calculations######################################
################################################################################

@@

#plotting simulated datapoints and smooth term
p14<-ggplot(data=dat,
            aes(x=time,y=y))+
  geom_point(size=sz+1)+
  scale_color_aaas()+
  labs(y='Simulated \n response')+
  geom_line(data=smooth,
            aes(x=time,
                y=smooth),
            color="#6C581DFF",
            size=sz+1)+
  theme_classic()
\end{lstlisting}

\begin{lstlisting}
## Error in scale_color_aaas(): could not find function "scale_color_aaas"
\end{lstlisting}

\begin{lstlisting}[language=R]
#Combining all
b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
  theme(
     text=element_text(size=18)
     )
\end{lstlisting}

@@

\begin{lstlisting}[language=R]
#see https://patchwork.data-imaginist.com/reference/area.html
layout1 <- c(
  area(1, 1),
  area( 1, 2),
  area(2, 1),
  area(2, 2),
  area(1, 3, 2)
)

layout2 <- c(
  area(1, 1),
  area( 1, 2),
  area(2, 1),
  area(2, 2),
  area(1,3,2,5)
)

#plot(layout2)
\end{lstlisting}

@@

\begin{lstlisting}
## 
## Method: REML   Optimizer: outer newton
## full convergence after 6 iterations.
## Gradient range [-4.297642e-08,-3.237913e-09]
## (score 437.636 & scale 390.173).
## Hessian positive definite, eigenvalue range [0.1251014,49.00133].
## Model rank =  5 / 5 
## 
## Basis dimension (k) checking results. Low p-value (k-index<1) may
## indicate that k is too low, especially if edf is close to k'.
## 
##          k'  edf k-index p-value    
## s(Day) 4.00 1.51    0.36  <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{lstlisting}

@@

\begin{lstlisting}
## 
## Family: gaussian 
## Link function: identity 
## 
## Formula:
## StO2_sim ~ s(Day, k = 5, bs = "gp")
## 
## Parametric coefficients:
##             Estimate Std. Error t value Pr(>|t|)    
## (Intercept)   20.111      1.975   10.18   <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Approximate significance of smooth terms:
##          edf Ref.df     F p-value   
## s(Day) 1.511  1.782 8.116 0.00349 **
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## R-sq.(adj) =  0.106   Deviance explained = 11.9%
## -REML = 437.64  Scale est. = 390.17    n = 100
\end{lstlisting}

@@

\begin{lstlisting}
## 
## Method: REML   Optimizer: outer newton
## full convergence after 9 iterations.
## Gradient range [-0.0001383864,0.0002277102]
## (score 414.2119 & scale 246.4112).
## Hessian positive definite, eigenvalue range [0.0001383659,48.50328].
## Model rank =  9 / 9 
## 
## Basis dimension (k) checking results. Low p-value (k-index<1) may
## indicate that k is too low, especially if edf is close to k'.
## 
##                         k'  edf k-index p-value    
## s(Day):GroupControl   4.00 1.00    0.47  <2e-16 ***
## s(Day):GroupTreatment 4.00 1.82    0.47  <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{lstlisting}

@@

\begin{lstlisting}
## 
## Family: gaussian 
## Link function: identity 
## 
## Formula:
## StO2_sim ~ s(Day, by = Group, k = 5, bs = "gp")
## 
## Parametric coefficients:
##             Estimate Std. Error t value Pr(>|t|)    
## (Intercept)    20.11       1.57   12.81   <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Approximate significance of smooth terms:
##                         edf Ref.df      F p-value    
## s(Day):GroupControl   1.001  1.001  5.243  0.0242 *  
## s(Day):GroupTreatment 1.823  2.071 35.305  <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## R-sq.(adj) =  0.435   Deviance explained = 45.1%
## -REML = 414.21  Scale est. = 246.41    n = 100
\end{lstlisting}

@@

m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
            method='REML',
            data  = dat_sim)

gam.check(m1)
\end{lstlisting}

@@

\begin{lstlisting}
## 
## Method: REML   Optimizer: outer newton
## full convergence after 7 iterations.
## Gradient range [-2.60407e-05,5.064354e-06]
## (score 368.6968 & scale 78.46844).
## Hessian positive definite, eigenvalue range [0.4967554,48.06648].
## Model rank =  10 / 10 
## 
## Basis dimension (k) checking results. Low p-value (k-index<1) may
## indicate that k is too low, especially if edf is close to k'.
## 
##                         k'  edf k-index p-value
## s(Day):GroupControl   4.00 3.19    1.17    0.94
## s(Day):GroupTreatment 4.00 3.79    1.17    0.95
\end{lstlisting}

\begin{lstlisting}[language=R]
summary(m1)
\end{lstlisting}

\begin{lstlisting}
## 
## Family: gaussian 
## Link function: identity 
## 
## Formula:
## StO2_sim ~ Group + s(Day, by = Group, k = 5, bs = "gp")
## 
## Parametric coefficients:
##                Estimate Std. Error t value Pr(>|t|)    
## (Intercept)       8.964      1.253   7.156 2.06e-10 ***
## GroupTreatment   22.293      1.772  12.583  < 2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Approximate significance of smooth terms:
##                         edf Ref.df     F  p-value    
## s(Day):GroupControl   3.191  3.437 11.05 1.01e-06 ***
## s(Day):GroupTreatment 3.788  3.957 63.23  < 2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## R-sq.(adj) =   0.82   Deviance explained = 83.5%
## -REML =  368.7  Scale est. = 78.468    n = 100
\end{lstlisting}

The resulting model is\passthrough{\lstinline!m1!}, which is the model fitted in the main manuscript. By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group . From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\textasciitilde1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 87\% of the variance data.

@@

\begin{lstlisting}
##               df      AIC
## gam_00  3.781952 885.4678
## gam_01  5.072091 840.7331
## gam1   10.936093 732.3416
\end{lstlisting}

@@

##matrix that contains the basis functions evaluated at the points in pdat
    xp <- predict(m1, newdata = pdat, type = 'lpmatrix')

@@

\begin{lstlisting}[language=R]
#missing data
#create a sequence of 40 random numbers between 1 and 100, these numbers will
#correspond to the row numbers to be randomly erased from the original dataset
set.seed(1)
missing <- sample(1:100, 40)

@@

\begin{lstlisting}
## # A tibble: 10 x 3
## # Groups:   Day, Group [10]
##      Day Group         n
##    <dbl> <fct>     <int>
##  1     0 Control       5
##  2     0 Treatment     3
##  3     2 Control       5
##  4     2 Treatment     5
##  5     5 Control       4
##  6     5 Treatment     6
##  7     7 Control       2
##  8     7 Treatment     7
##  9    10 Control       1
## 10    10 Treatment     6
\end{lstlisting}

### 2021-05-28 Manuscripts/Manuscript by chapters -v2/02-Challenges.Rmd
Updated, added LMEM and rm-ANOVA models in text and in appendix, corrected cubic spline to thin plate spline, changed to default in gam models instead of gp smooths

To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each. An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme]. 
Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A). In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response), C). In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point. This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments.

@@

A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal appears in Figure \@ref(fig:linear-cases-Appendix) and Figure \@ref(fig:quadratic-cases-Appendix) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented. We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs. Of notice, Section \@ref(longitudinal-GAMs) uses simulated data that does follow reported trends in the biomedical literature.

@@

   #plot the model predictions
  p4 <- ggplot(sim_dat$dat, 
               aes(x = time, 
                   y = y, 
                   color = treatment)) +
    geom_point()+
    labs(y='response')+
    geom_line(aes(y = predict(sim_dat$fit_lme), 
                  group = subject, size = "Subjects")) +
    geom_line(data = sim_dat$pred_dat, 
              aes(y = predict(sim_dat$fit_lme, 
                              level = 0, 
                              newdata = sim_dat$pred_dat), 
                  size = "Population")) +
    guides(color = guide_legend(override.aes = list(size = 2)))+
    scale_size_manual(name = "Predictions", 
                      values=c("Subjects" = 0.5, "Population" = 3)) +
    theme_classic() +
    theme(plot.title = element_text(size = 30, 
                                  face = "bold"),
        text=element_text(size=30))+
    thm

  return((p1+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))

@@

(ref:l-q-response-caption) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model. A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data. B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic). Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data. The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.

@@

The simulation shows that the fit produced by the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response),B). When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.

However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:l-q-response), C. Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in (Figure \@ref(fig:l-q-response) A and B) . The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group. Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears in panel D in Figure \@ref(fig:l-q-response).

A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\@ref(fig:l-q-response), D) indicates that the model is not capturing the changes within each group.  Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model. Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels C and D). If such a change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response. Thus, even though the model correctly detects a difference in treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.

>>>>

To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each. An rm-ANOVA model (Equation \@ref(eq:linear-model)), and a LMEM  (Equation \@ref(eq:LMEM)) are fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme].

Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A). In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response), C). In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point. This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments.

@@

A more comprehensive exploration  of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal appears in Figure \@ref(fig:linear-cases-Appendix) and Figure \@ref(fig:quadratic-cases-Appendix) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented. We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs. Of notice, Section \@ref(longitudinal-GAMs) uses simulated data that does follow reported trends in the biomedical literature.

@@

   #plot the model predictions for rm-ANOVA
  p2 <- ggplot(sim_dat$dat, 
               aes(x = time, 
                   y = y, 
                   color = treatment)) +
    geom_point(show.legend=FALSE)+
    labs(y='response')+
    geom_line(aes(y = predict(sim_dat$fit_anova), 
                  group = subject, size = "Subjects"),show.legend = FALSE) +
    geom_line(data = sim_dat$pred_dat, 
              aes(y = predict(sim_dat$fit_anova, 
                              level = 0, 
                              newdata = sim_dat$pred_dat), 
                  size = "Population"),
              show.legend=FALSE) +
    guides(color = guide_legend(override.aes = list(size = 2)))+
    scale_size_manual(name = "Predictions", 
                      values=c("Subjects" = 0.5, "Population" = 3)) +
    theme_classic() +
    theme(plot.title = element_text(size = 30, 
                                  face = "bold"),
        text=element_text(size=30))+
    thm

   #plot the model predictions for LMEM
  p4 <- ggplot(sim_dat$dat, 
               aes(x = time, 
                   y = y, 
                   color = treatment)) +
    geom_point()+
    labs(y='response')+
    geom_line(aes(y = predict(sim_dat$fit_lme), 
                  group = subject, size = "Subjects")) +
    geom_line(data = sim_dat$pred_dat, 
              aes(y = predict(sim_dat$fit_lme, 
                              level = 0, 
                              newdata = sim_dat$pred_dat), 
                  size = "Population")) +
    guides(color = guide_legend(override.aes = list(size = 2)))+
    scale_size_manual(name = "Predictions", 
                      values=c("Subjects" = 0.5, "Population" = 3)) +
    theme_classic() +
    theme(plot.title = element_text(size = 30, 
                                  face = "bold"),
        text=element_text(size=30))+
    thm

  return((p1+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))

@@

(ref:l-q-response-caption) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a LMEM and a rm-ANOVA model. A, D: Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data. B,E: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic). Points represent the original raw data. The rm-ANOVA model not only fails to pick the trend of the quadratic data but also assigns a global estimate that does not take between-subject variation. C, F: Estimates from the LMEM model in the linear and quadratic case.The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data in each group and grossly bias the initial estimates for each group.

@@

The simulation shows that the fit produced by the LMEM and the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response),B, E). When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.

However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:l-q-response), C. Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in (Figure \@ref(fig:l-q-response) A and B) . The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group. Fitting an rm-ANOVA model \@ref(eq:linear-model) or a LMEM \@ref(eq:LMEM) to this data  produces the fit that appears in panels E and F in Figure \@ref(fig:l-q-response).

A comparison of the fitted mean response of the LMEM and the rm-ANOVA model to the simulated data in Figure ((\@ref(fig:l-q-response), E, F) indicates that the models are not capturing the changes within each group.  Specifically, note that the fitted mean response of both models (panel E, F) show that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured. The LMEM is only able to account for between-subject variation by providing different intercepts to each subject, but both models are not able to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panel D with panels E and F). If such a change has important physiological implications,both rm-ANOVA and LMEMs omit it from the fitted mean response. Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.

### 2021-05-28 Manuscripts/Manuscript by chapters -v2/03-GAM Theory.Rmd
Updated, added LMEM and rm-ANOVA models in text and in appendix, corrected cubic spline to thin plate spline, changed to default in gam models instead of gp smooths

A commonly used _basis function_ is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness [@wood2017;@simpson2018]. Cubic splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983]. Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.

>>>>

Commonly used _basis functions_ are splines (cubic, thin plate regression among others). A cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data [@wood2017;@simpson2018]. Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983]. Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.

### 2021-05-28 Manuscripts/Manuscript by chapters -v2/04-Longitudinal analysis with GAMs.Rmd
Updated, added LMEM and rm-ANOVA models in text and in appendix, corrected cubic spline to thin plate spline, changed to default in gam models instead of gp smooths

`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5,bs="gp"), method='REML',data=dat_sim)`

This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots. The smooth is constructed using gaussian process smooths, which is indicated by `bs="gp"`. These splines are used to model temporal trends and might be particularly suited for long-term studies where the correlation between measurements changes as a function of the time intervals [@simpson2018]. The parametric term _Group_ is added to quantify differences in the effect of treatment between groups, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017]. When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),B). Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_ [@gratia]. A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].

@@

gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
            method='REML',
            data  = dat_sim)

>>>>

`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5), method='REML',data=dat_sim)`

This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots. The smooth is constructed by default using thin plate regression splines. Other splines can be used if desired, including gaussian process smooths [@simpson2018]. The parametric term _Group_ is added to quantify differences in the effect of treatment between groups, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017]. When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),B). Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_ [@gratia]. A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].

@@

gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
            method='REML',
            data  = dat_sim)
